% bibliography.tex
% Manual bibliography entries for the pamphlet (no BibTeX/BibLaTeX)

\section*{References and further reading}
\addcontentsline{toc}{section}{References and further reading}

\begin{singlespace}

\begin{thebibliography}{99}
\setlength{\itemsep}{0.35em}
\setlength{\parsep}{0pt}

\bibitem{Marx1867}
Marx, K. (1867) \emph{Capital: A Critique of Political Economy, Volume I}.
Available at: \url{https://www.marxists.org/archive/marx/works/1867-c1/}.

\bibitem{Marx1894}
Marx, K. (1894) \emph{Capital: A Critique of Political Economy, Volume III}.
(Edited by F. Engels.)
Available at: \url{https://www.marxists.org/archive/marx/works/1894-c3/}. % confirmed URL form

\bibitem{MarxGrundrisse}
Marx, K. (1857--58) \emph{Grundrisse: Foundations of the Critique of Political Economy}.
Available at: \url{https://www.marxists.org/archive/marx/works/1857/grundrisse/}.

\bibitem{EngelsAntiDuhring}
Engels, F. (1878) \emph{Anti-D\"uhring}.
Available at: \url{https://www.marxists.org/archive/marx/works/1877/anti-duhring/}.

\bibitem{LeninImperialism}
Lenin, V. I. (1916) \emph{Imperialism, the Highest Stage of Capitalism}.
Available at: \url{https://www.marxists.org/archive/lenin/works/1916/imp-hsc/}.

\bibitem{PatnaikImperialism}
Patnaik, U. and Patnaik, P. (2016) \emph{A Theory of Imperialism}.
New York: Columbia University Press.

\bibitem{Shaikh2016}
Shaikh, A. (2016) \emph{Capitalism: Competition, Conflict, Crises}.
Oxford: Oxford University Press.

\bibitem{Roberts2015}
Roberts, M. (2016) \emph{The Long Depression: Marxism and the Global Crisis of Capitalism}.
Chicago: Haymarket Books.

\bibitem{HarveyNewImperialism}
Harvey, D. (2003) \emph{The New Imperialism}.
Oxford: Oxford University Press.

\bibitem{HarveyCompanion}
Harvey, D. (2010) \emph{A Companion to Marx's Capital, Volume 1}.
London: Verso.

\bibitem{BhattacharyaSRT}
Bhattacharya, T. (ed.) (2017) \emph{Social Reproduction Theory: Remapping Class, Recentering Oppression}.
London: Pluto Press.
% :contentReference[oaicite:2]{index=2}

\bibitem{FedericiCaliban}
Federici, S. (2004) \emph{Caliban and the Witch: Women, the Body and Primitive Accumulation}.
Brooklyn: Autonomedia.
% :contentReference[oaicite:3]{index=3}

\bibitem{EngelsFuneral}
Engels, F. (1883) ``Speech at the Grave of Karl Marx''.
Available at: \url{https://www.marxists.org/archive/marx/works/1883/death/burial.htm}.

\bibitem{Smith2016}
Smith, J. (2016) \emph{Imperialism in the Twenty-First Century: Globalization, Super-Exploitation, and Capitalism's Final Crisis}.
New York: Monthly Review Press.

\bibitem{Hickel2022}
Hickel, J., Sullivan, D. and Zoomkawala, H. (2022) ``Plunder in the Post-Colonial Era: Quantifying Drain from the Global South through Unequal Exchange, 1990--2015''.
\emph{Global Environmental Change}, 73, 102467.
DOI: \url{https://doi.org/10.1016/j.gloenvcha.2022.102467}.
% :contentReference[oaicite:4]{index=4}

% --- Replace/upgrade the iPhone/iPod value-chain citation ---
\bibitem{DedrickKraemerLinden2010}
Dedrick, J., Kraemer, K. L. and Linden, G. (2010) ``Who profits from innovation in global value chains? A study of the iPod and notebook PCs''.
\emph{Industrial and Corporate Change}, 19(1), 81--116.
DOI: \url{https://doi.org/10.1093/icc/dtp032}.

% --- Algorithmic management / platform labour ---
\bibitem{WoodEtAl2019}
Wood, A. J., Graham, M., Lehdonvirta, V. and Hjorth, I. (2019) ``Good Gig, Bad Gig: Autonomy and Algorithmic Control in the Global Gig Economy''.
\emph{Work, Employment and Society}, 33(1), 56--75.
DOI: \url{https://doi.org/10.1177/0950017018785616}.

% --- Data centres: energy system baseline source (fill with the exact IEA page/report you intend to cite) ---
\bibitem{IEADataCentres2019}
International Energy Agency (IEA) (2019) \emph{Data Centres and Data Transmission Networks}.
Paris: IEA. Available at: \url{https://www.iea.org/}.

% --- Green capitalism anchor ---
\bibitem{IPCCAR6WGIII}
IPCC (2022) \emph{Climate Change 2022: Mitigation of Climate Change}.
Contribution of Working Group III to the Sixth Assessment Report of the Intergovernmental Panel on Climate Change.
Cambridge: Cambridge University Press.

% --- Neoliberalism / post-1970s turn ---
\bibitem{HarveyNeoliberalism}
Harvey, D. (2005) \emph{A Brief History of Neoliberalism}.
Oxford: Oxford University Press.

% --- Structural adjustment / IMF-WB-WTO political economy anchor ---
\bibitem{PeetUnholyTrinity}
Peet, R. (2003) \emph{Unholy Trinity: The IMF, World Bank and WTO}.
London: Zed Books.

\bibitem{TrotskyPermanentRevolution}
Trotsky, L. (1931) \emph{The Permanent Revolution}.
Available at: \url{https://www.marxists.org/archive/trotsky/1931/tpr/}.

\bibitem{RosenbergUCD}
Rosenberg, J. (2006) ``Why is there no International Historical Sociology?'' \emph{European Journal of International Relations}, 12(3), 307--340.
(Develops the concept of uneven and combined development in IR/IHS.)

\bibitem{StrugglePK_TDR}
The Struggle (Pakistan) (2024) \emph{Pakistan Perspective 2024--25} [PDF].
Available at: \url{https://www.struggle.pk/wp-content/uploads/2024/02/The-Struggle-Pakistan-Perspective-2024-25-PDF-Final.pdf}.

\bibitem{StrugglePK_WorldPerspectives}
Asian Marxist Review (2025) \emph{A Turning Point in History: Trump, Crisis, and the Decline of the Liberal Order} (World Perspectives 2025) [English translation of the first draft of the Congress 2025 document].
Available at: \url{https://www.marxistreview.asia/wp-2025/}.

\end{thebibliography}

\end{singlespace}
